The pragmatics of perspective-taking

### 2018-03-22 writing/journal_manuscript/manuscript.tex
fiddle with wording

This suggests that we consider whether the utterances produced by the confederate in Keysar's critical conditions were actually what a speaker in that context would be expected to say. 
If not, then perhaps listeners are making choices that are in fact consistent with a correct pragmatic interpretation of the confederate's (uncooperative) utterance.
More precisely, when both players know that objects are occluded in the display, the speaker may tend to add additional precision to references in order to avoid confusion. If the listener \emph{expects} the speaker to do this, they will pragmatically pick the \emph{a priori} more likely referent of the referring expression, which in critical trials will be the occluded object. In other words, it is precisely \emph{because} the listener takes the speaker's mental state into consideration that they are tricked by an uncooperative confederate into choosing the wrong item. 
%\begin{table*}
%\begin{center}
%\begin{tabular}{ p{2cm} | r | r |  r || r | r | r || r | r | r}
%& \multicolumn{3}{c||}{\% attempted at least once} & \multicolumn{3}{c||}{\% attempted at least twice} & \multicolumn{3}{c}{\% of total cases}\\
%\hline
%& Orig. & Expt. 1 & Expt. 2 & Orig. & Expt. 1 & Expt. 2 & Orig. & Expt. 1 & Expt. 2  \\
%Experimental & 71 & 93 & 61 & 46 & 57 & 32 & 30 & 43 & 24\\
%Baseline        & 0   & 7   & 0   & 0   & 0   & 0   & 0   & 2   & 0\\
%\end{tabular}
%\caption{Side-by-side comparison of error rates in \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} and our two replications. The first two sections show the percentage of \emph{participants} attempting to move the occluded distractor at least once, or twice, of the four possible cases. The third section shows the percentage of \emph{all experimental trials} that the participant actually tried to move the occluded object. }
%\label{table:mainResults}
%\end{center}
%\end{table*}
%
%We began by replicating \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} in a multi-player web experiment. We recruited participants to be both director and matcher (instead of using a confederate), but instructions for critical items, as well as a random subset of filler items, remained scripted as in the original study. We replicated the original finding, but noted a tendency of directors to be overinformative in unscripted filler trials. We then ran the same experiment without using any scripted instructions, observing unconstrained director utterances. We found much greater precision in unconstrained director utterances, which match targets much better than distractors, and better performance of the matchers. This minimal pair of experiments demonstrates that listener mistakes are at least partially due to the pragmatics of the task, ironically showing that apparent failures of theory of mind are in fact attributable to sophisticated expectations about speaker behavior---that is, to theory of mind.

In particular, we argue that critical utterances used by Keysar and colleagues are \emph{uncooperative}: they are less informative than what a speaker would actually produce in that situation; listeners who are sensitive to the pragmatics of the situation expect these more informative utterances and produce ``errors'' when their expectations are flouted. In Experiment 1, we conduct a direct replication of \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}, manipulating whether the speaker is a confederate or naive participant. We find that unscripted speakers naturally produce more informative utterances than those in the confederate's script. In Experiment 2, we test the hypothesis that speakers systematically produce longer, more specific utterances than strictly required by their visible context when they know objects are occluded. Finally, in Experiment 3, we manipulate the listener's pragmatic beliefs about whether the speaker is indeed expected to produce such `over-informative' utterances. Taken together, these results suggest that apparent failures of perspective-taking are in fact driven by sophisticated pragmatic reasoning; in other words, by successful theory of mind use.

@@

%To deal with participants arriving at unpredictable times, we implemented a waiting room mechanism, where the first person to enter the game environment would see a screen saying ``Waiting for another player..." and the second person would trigger the game to start. 
Participants were randomly assigned to `director' and `matcher' roles and placed in a shared environment containing a chat box and a $4 \times 4$ grid of objects (see Fig. \ref{fig:interface}). Five of the grid cells were blocked from the director's view by a curtain, and six or seven objects were displayed in the grid at a given time. Trials were blocked into eight `items', each using a different set of objects. For each item, we gave the director a series of four instructions to move objects, which were displayed as a series of arrows pointing from some object to a neighboring unoccupied cell. They used a chat box to communicate the identity of this object and action to their partner, who proceeded to click and drag an object to complete the trial.

We used a between-subject design to test whether the scripted confederate labels were in fact what cooperative speakers would say in context. In the `scripted' condition, directors were free to communicate however they wished for half of the instructions but for the other half, including all the critical instructions, their messages were pre-scripted using the precise wording from \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}. For example, when giving instructions on how to move the cassette tape, the director would be forced to use the ambiguous utterance ``Move the tape down one square." (That is, in these conditions, the scripted message would automatically appear in the director's chat box, and they would have to click `send' for the experiment to continue.) In the `unscripted' condition, directors were unrestricted and free to send whatever messages they deemed appropriate on all trials.

In each block, one target belonged to a `critical pair' of objects, such as a visible cassette tape and a hidden roll of tape that could both plausibly be referred to as `the tape'. We collected baseline performance for each condition by replacing the hidden alternative (e.g.~a roll of tape) with an object that did not fit the critical instruction (e.g.~a battery); we used the same unambiguous replacements as \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}. Each participant received half the items in the experimental condition and half in the baseline condition. The assignment of items to conditions was randomized across participants, and the order of conditions was randomized under the constraint that the same condition would not be used on more than two consecutive items. All object sets, object placements, and corresponding instruction sets were the same for all participants.

@@

To collect clean mouse-tracking data, we asked the matcher to wait until the director sent a message; when the message was received, the matcher clicked a small circle in the center of the grid to continue the trial. We started recording from the matcher's mouse after this click.

@@

As a proxy for the eye-tracking analyses reported by \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}, we conducted a mouse-tracking analysis. Mouse-position is highly correlated with gaze in web browsing \citep{ChenEtAl01_MouseCursor, SpiveyGrosjeanKnoblich05} and has been used  %We define the decision window as the span of time between  their instruction message and when they started moving an object. If it took them multiple attempts to move the correct object, we restricted our analysis to the first attempt. 
Within the decision window, we computed the total proportion of time spent moving toward the target relative to the distractor. %We had to exclude an additional 3 participants for this analysis, because the timestamps for director and matcher did not align and we could not establish the decision window properly. 
%A paired-samples $t$-test found that matchers in the scripted condition tended to spend less time hovering over the target cell on critical experimental trials than on baseline trials, $t(10) = -2.65, p = 0.02$ (see Fig. \ref{fig:hover}), indicating that the presence of a hidden distractor interfered with participants ability to directly choose the target.%\footnote{This analysis includes participants who actually made errors, since the data is too sparse to exclude them.} . %accounts for a significant proportion of variance $F(3, 52) = 10.7, p < 0.001$

@@

By running this replication as a multi-player web experiment we have available an additional source of data beyond the original experiments: half of the instructions were unscripted, providing observations of natural production of referential descriptions for filler items. 
Informally, we noted a tendency toward additional, possibly unnecessary, precision in descriptions. Instead of ``move the stuffed animal down'', participants said ``move the stuffed panda bear down.'' Or, instead of saying ``move the plane to the right'' when there is only one plane, participants said ``move the red airplane to the right.'' 
Perhaps directors were taking the time to make more precise descriptions because they believed it was contextually relevant: both parties know that there are hidden objects in the environment increasing the chance of miscommunication from imprecise descriptions. 
If the matcher expected the director to be precise, then they would be justified in picking the first or best object that meets the description (rather than worrying excessively about occluded cells). 
That is, the scripted instructions used by the director for critical trials may have been \emph{uncooperative} for this situation, and thus led matchers astray. 
We expected to see more precise descriptions by unscripted directors and fewer errors by matchers in the critical trials.

@@

Still, in running this replication as a multi-player web experiment, we observed two particularly interesting features in the data. First,

@@

  \begin{figure}[t!]
 \begin{center}
 \vspace{-.5cm}
 \includegraphics[scale=.3]{figures/Exp2_design}
 \caption{Within-pair factorial design used in Exp.~2. In the common condition, we reproduce a simple pragmatic reference contrast: when more than one star is present in common ground, speakers are expected to be more informative. On hidden versions of these condition, the same set of objects is visible, but some additional squares are occluded. If pragmatic directors represent uncertainty over what the matcher sees behind these squares, we predict that they will produce more informative utterances.}
 \vspace{-.5cm}
 \label{fig:hover}
 \end{center}
 \end{figure}

@@

\item A pragmatic listener, reasoning about this uncertain speaker, is presented with a full array of four objects $\mathcal{O}_L$, one of which is identified as hidden to the speaker, $o_h$. The listener, however, also has some uncertainty over whether the experimenter is deceptive, represented by a Bernoulli random variable $d$, such that they assign a small prior probability $P(d)$ to the possibility that $o_h$ is actually in common ground. Upon hearing an utterance$u$, they update both their beliefs about which object is the target and whether the experimenter is being deceptive:
$$L_2(t, d | u, \mathcal{O}_L) = P(t)\left(S(u | t, \mathcal{O}_L)P(d)\right)^d(S(u | t, \mathcal{O}_L - o_h)(1-P(d))^{1-d}$$
\end{enumerate}
Technical details are omitted here for brevity, but the full model is fully specified in WebPPL and available to run at %\url{http://forestdb.org/models/keysar.html}.

>>>>

This suggests that we consider whether the scripted utterances produced by  confederates in popular perspective-taking tasks were actually what a speaker in that context would be expected to say \citep{KuhlenBrennan13_LanguageInDialogue}. 
If not, then perhaps listeners are making choices that are in fact consistent with a correct pragmatic interpretation of the confederate's (uncooperative) utterance.
More precisely, when both players know that objects are occluded in the display, speakers add additional precision to references in order to avoid confusion. 
If the listener justifiably expects the speaker to do so, they may \emph{rationally} adopt a strategy of choosing the \emph{a priori} more likely referent of the referring expression regardless of perspective. 
In other words, it is precisely \emph{because} the listener takes the speaker's mental state into consideration that they are tricked by an uncooperative confederate into choosing the wrong item. 
%\begin{table*}
%\begin{center}
%\begin{tabular}{ p{2cm} | r | r |  r || r | r | r || r | r | r}
%& \multicolumn{3}{c||}{\% attempted at least once} & \multicolumn{3}{c||}{\% attempted at least twice} & \multicolumn{3}{c}{\% of total cases}\\
%\hline
%& Orig. & Expt. 1 & Expt. 2 & Orig. & Expt. 1 & Expt. 2 & Orig. & Expt. 1 & Expt. 2  \\
%Experimental & 71 & 93 & 61 & 46 & 57 & 32 & 30 & 43 & 24\\
%Baseline        & 0   & 7   & 0   & 0   & 0   & 0   & 0   & 2   & 0\\
%\end{tabular}
%\caption{Side-by-side comparison of error rates in \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} and our two replications. The first two sections show the percentage of \emph{participants} attempting to move the occluded distractor at least once, or twice, of the four possible cases. The third section shows the percentage of \emph{all experimental trials} that the participant actually tried to move the occluded object. }
%\label{table:mainResults}
%\end{center}
%\end{table*}
%
%We began by replicating \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} in a multi-player web experiment. We recruited participants to be both director and matcher (instead of using a confederate), but instructions for critical items, as well as a random subset of filler items, remained scripted as in the original study. We replicated the original finding, but noted a tendency of directors to be overinformative in unscripted filler trials. We then ran the same experiment without using any scripted instructions, observing unconstrained director utterances. We found much greater precision in unconstrained director utterances, which match targets much better than distractors, and better performance of the matchers. This minimal pair of experiments demonstrates that listener mistakes are at least partially due to the pragmatics of the task, ironically showing that apparent failures of theory of mind are in fact attributable to sophisticated expectations about speaker behavior---that is, to theory of mind.

%In particular, we argue that critical utterances used by Keysar and colleagues are \emph{uncooperative}: they are less informative than what a speaker would actually produce in that situation; listeners who are sensitive to the pragmatics of the situation expect these more informative utterances and produce ``errors'' when their expectations are flouted.
In Experiment 1, we conduct a direct replication of \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}, manipulating whether the speaker is a confederate or naive participant. We find that unscripted speakers naturally produce more informative utterances than those in the confederate's script. In Experiment 2, we test the hypothesis that speakers systematically produce longer, more specific utterances than strictly required by their visible context when they know objects are occluded. %Finally, in Experiment 3, we manipulate the listener's pragmatic beliefs about whether the speaker is indeed expected to produce such `over-informative' utterances. 
Taken together, these results suggest that apparent failures of perspective-taking are in fact attributable to sophisticated pragmatic expectations about speaker behavior---that is, to successful theory of mind use.

 \begin{figure}[t!]
 \begin{center}
 \vspace{-.25cm}
 \includegraphics[scale=.5]{figures/Exp1_design.png}
 \vspace{-.5cm}
 \caption{Screenshots of interface on a critical trial: a cassette tape is in view of both players, but a roll of tape is occluded from the speaker's view.}
 \vspace{-.5cm}
 \label{fig:exp1paradigm}
 \end{center}
 \end{figure}

@@

%To deal with participants arriving at unpredictable times, we implemented a waiting room mechanism, where the first person to enter the game environment would see a screen saying ``Waiting for another player..." and the second person would trigger the game to start. 
Participants were randomly assigned to `director' and `matcher' roles and placed in a shared environment containing a chat box and a $4 \times 4$ grid of objects (see Fig. \ref{fig:exp1paradigm}). Trials were blocked into eight `items', each using a different set of objects. For each item, we gave the director a series of four instructions, which were displayed as a series of arrows pointing from some object to a neighboring unoccupied cell. They used a chat box to communicate the identity of this object and action to their partner, who proceeded to click and drag an object to complete the trial.

We used a between-subject design to test whether the scripted confederate labels were in fact what cooperative speakers would say in context. Directors assigned to the `scripted' condition were free to communicate however they wished for half of the instructions. For the other half, including all the critical instructions, their messages were pre-scripted using the precise wording from \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}. For example, when giving instructions on how to move the cassette tape, the director would be forced to use the ambiguous utterance ``Move the tape down one square." (That is, in these conditions, the scripted message would automatically appear in the director's chat box, and they would have to click `send' for the experiment to continue.) In the `unscripted' condition, directors were unrestricted and free to send whatever messages they deemed appropriate on all trials.

In each block, one target belonged to a `critical pair' of objects, such as a visible cassette tape and a hidden roll of tape that could both plausibly be referred to as `the tape'. We collected baseline performance for each condition by replacing the hidden alternative (e.g.~a roll of tape) with an object that did not fit the critical instruction (e.g.~a battery); we used the same unambiguous replacements as \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}. Each participant received half the items in the experimental condition and half in the baseline condition. The assignment of items to conditions was randomized across participants, and the order of conditions was randomized under the constraint that the same condition would not be used on more than two consecutive items. Five of the grid cells were blocked from the director's view by a curtain, and six or seven objects were displayed in the grid at a given time. All object sets, object placements, and corresponding instruction sets were fixed across participants.

In addition to collecting messages sent through the chat box and matcher errors, we collected mouse-tracking data in analogy to the eye-tracking common in these paradigms. To collect clean data, we asked the matcher to wait until the director sent a message; when the message was received, the matcher clicked a small circle in the center of the grid to show the objects and proceed with the trial. We recorded at 100Hz from the matcher's mouse in the decision window after this click.

@@

@@

As a proxy for the eye-tracking analyses reported by \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}, we conducted a mouse-tracking analysis. Mouse-position is highly correlated with gaze in web browsing \citep{ChenEtAl01_MouseCursor, SpiveyGrosjeanKnoblich05} and has been used  %We define the decision window as the span of time between  their instruction message and when they started moving an object. If it took them multiple attempts to move the correct object, we restricted our analysis to the first attempt. 
% Within the decision window, we computed the total proportion of time spent moving toward the target relative to the distractor. %We had to exclude an additional 3 participants for this analysis, because the timestamps for director and matcher did not align and we could not establish the decision window properly. 
%A paired-samples $t$-test found that matchers in the scripted condition tended to spend less time hovering over the target cell on critical experimental trials than on baseline trials, $t(10) = -2.65, p = 0.02$ (see Fig. \ref{fig:hover}), indicating that the presence of a hidden distractor interfered with participants ability to directly choose the target.%\footnote{This analysis includes participants who actually made errors, since the data is too sparse to exclude them.} . %accounts for a significant proportion of variance $F(3, 52) = 10.7, p < 0.001$

@@

By running this replication as a multi-player web experiment we have available an additional source of data beyond the original experiments: half of the instructions were unscripted, providing observations of natural production of referential descriptions for filler items. 
Informally, we noted a tendency toward additional, possibly unnecessary, precision in descriptions. Instead of ``move the stuffed animal down'', participants said ``move the stuffed panda bear down.'' Or, instead of saying ``move the plane to the right'' when there is only one plane, participants said ``move the red airplane to the right.'' 
Perhaps directors were taking the time to make more precise descriptions because they believed it was contextually relevant: both parties know that there are hidden objects in the environment increasing the chance of miscommunication from imprecise descriptions. 
%If the matcher expected the director to be precise, then they would be justified in picking the first or best object that meets the description (rather than worrying excessively about occluded cells). 
%That is, the scripted instructions used by the director for critical trials may have been \emph{uncooperative} for this situation, and thus led matchers astray. 
We therefore expected to see more precise descriptions by unscripted directors and fewer errors by matchers in the critical trials.

@@

%Still, in running this replication as a multi-player web experiment, we observed two particularly interesting features in the data. First,

@@

  \begin{figure}[t!]
 \begin{center}
 \vspace{-.5cm}
 \includegraphics[scale=.25]{figures/Exp2_design}
 \caption{Within-pair factorial design used in Exp.~2. In the common condition, we reproduce a simple pragmatic reference contrast: when more than one star is present in common ground, speakers are expected to be more informative. On hidden versions of these condition, the same set of objects is visible, but some additional squares are occluded. If pragmatic directors represent uncertainty over what the matcher sees behind these squares, we predict that they will produce more informative utterances.}
 \vspace{-.5cm}
 \label{fig:hover}
 \end{center}
 \end{figure}

@@

%\item A pragmatic listener, reasoning about this uncertain speaker, is presented with a full array of four objects $\mathcal{O}_L$, one of which is identified as hidden to the speaker, $o_h$. The listener, however, also has some uncertainty over whether the experimenter is deceptive, represented by a Bernoulli random variable $d$, such that they assign a small prior probability $P(d)$ to the possibility that $o_h$ is actually in common ground. Upon hearing an utterance$u$, they update both their beliefs about which object is the target and whether the experimenter is being deceptive:
%$$L_2(t, d | u, \mathcal{O}_L) = P(t)\left(S(u | t, \mathcal{O}_L)P(d)\right)^d(S(u | t, \mathcal{O}_L - o_h)(1-P(d))^{1-d}$$
\end{enumerate}
Technical details are omitted here for brevity, but the full model is fully specified in WebPPL and available to run at %\url{http://forestdb.org/models/keysar.html}.
